%% AAAI-27 anonymous submission — single-tool-use evaluation paper.
%% SCAFFOLD ONLY: sections are empty (TODO pointers, no prose). See GOALS.md for scope.
%% Full formatting rules: authorkit27/AnonymousSubmission2027.pdf
%% Preamble below is the AAAI-27 kit preamble verbatim — DO NOT CHANGE the marked lines.

\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai2027}  % DO NOT CHANGE THIS  (anonymous submission mode)
% Fonts are loaded automatically by aaai2027.sty (newtxtext, helvet, courier).
% DO NOT add \usepackage{times}, {helvet}, {courier}, or any other font package.
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS

% Recommended for algorithms (remove if unused before final submission).
\usepackage{algorithm}
\usepackage{algorithmic}

% Recommended for listings, e.g. PDDL snippets (remove if unused before final submission).
\usepackage{newfloat}
\usepackage{listings}
\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
\lstset{%
	basicstyle={\footnotesize\ttfamily},
	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,
	aboveskip=0pt,belowskip=0pt,
	showstringspaces=false,tabsize=2,breaklines=true}
\floatstyle{ruled}
\newfloat{listing}{tb}{lst}{}
\floatname{listing}{Listing}

% Recommended for better-looking tables.
\usepackage{booktabs}

\pdfinfo{
/TemplateVersion (2027.1)
}

\setcounter{secnumdepth}{0} % 0 = unnumbered sections (AAAI default). Set to 1/2 to number.

% ============================ Title & Authors ============================
% Title is NOT anonymized (only authors are). Title Case (Chicago) required.
\title{Working Title (TBD)} % TODO: set the real title before submission.

% Anonymous submission: sole "author" is "Anonymous Submission"; affiliations empty.
% De-anonymize (real authors/affiliations) only at camera-ready via CameraReady2027.tex.
\author{
    Anonymous Submission
}
\affiliations{
    %
}

\begin{document}

\maketitle

\begin{abstract}
% TODO (goal 1): Abstract. Write LAST. ~150-200 words, no citations.
\end{abstract}

% Optional anonymized links block (uncomment near submission; do NOT de-anonymize).
% \begin{links}
%     \link{Code}{...}
%     \link{Datasets}{...}
% \end{links}

\section{Introduction}
% TODO: motivation; the gap (do symbolic PDDL tools actually help LLMs?); our framing;
% explicit contributions list. Self-cite the earlier arXiv version anonymously
% (see refs.bib note) — it is our own prior work.

\section{Background and Related Work}
% DRAFT 2026-06-14 (goal 2). Re-derived for our framework; cites the earlier version
% (Benyamin et al.) in the THIRD PERSON only. Some refs flagged [VERIFY] in refs.bib.

Automated planning studies how to synthesize a course of action that transforms an
initial state into one satisfying a goal. The Planning Domain Definition Language (PDDL)
\citep{mcdermott1998pddl, fox2003pddl21} is the field's standard formalism, and decades of
research have produced sound, complete solvers such as Fast Downward
\citep{helmert2006fastdownward} and FF \citep{hoffmann2001ff} for classical planning and
ENHSP \citep{scala2016enhsp} for numeric planning, together with plan validators such as
VAL \citep{howey2004val}. These systems are correct by construction: a plan they return
provably reaches the goal, and a verdict they issue is sound. They are, however, brittle to
the formal-modelling step --- a domain or problem must be expressed in precise PDDL before
any solver can run.

\subsection{Large Language Models for Planning}
Large language models have been probed extensively as planners. Systematic benchmarks show
that, prompted to produce plans directly, even strong LLMs fall well short of solver-level
reliability \citep{valmeekam2023planbench}. \citet{kambhampati2024llmmodulo} argue from this
evidence that autoregressive LLMs cannot plan or self-verify on their own, and propose
\emph{LLM-Modulo} frameworks in which the model is paired with external, model-based
verifiers that supply the soundness it lacks. The present study is empirical evidence for
exactly this division of labour: we measure what an LLM gains when a sound planner or
validator is available as a callable tool, rather than asking the model to be the planner.

\subsection{Large Language Models as Planning Formalizers}
A complementary line treats the LLM not as the planner but as a \emph{formalizer} that
translates natural language into PDDL, delegating search to a sound solver. LLM+P
\citep{liu2023llmp} pipelines an LLM into a classical planner; \citet{guan2023world} use
LLMs to construct world models for model-based planning; Plansformer
\citep{pallagani2022plansformer} fine-tunes for plan generation; and pretrained LLMs have
been applied to generalized planning in PDDL domains \citep{silver2024generalized}.
\citet{tantakoun2025survey} survey this rapidly growing ``LLMs-as-formalizers'' area, and
\citet{huang2025limit} characterize its limits, showing that formalization quality degrades
sharply on obfuscated domains. This literature largely studies the \emph{generation} of
PDDL, evaluated on a single downstream question (does the produced model solve?). We instead
evaluate LLMs \emph{using} planning and validation tools across five distinct PDDL tasks,
with strict end-to-end grading against a solver oracle.

\subsection{Tool Use, Agent Protocols, and Their Evaluation}
Equipping LLMs with external tools is now standard, from learned tool invocation
\citep{schick2023toolformer} to large API repertoires \citep{qin2024toolllm}. The Model
Context Protocol (MCP) \citep{anthropic2024mcp} standardizes how tools are described and
called, and is the mechanism by which we expose planners and validators to the model. A
growing body of benchmarks measures tool-use competence --- single-turn function calling
\citep{bfcl2025} and multi-turn tool-agent-user interaction \citep{yao2024taubench}, among
others. Our setting sits deliberately at the simple end of this spectrum: each task admits
exactly one relevant, correct-by-construction tool, so the appropriate analog is single-call
function selection rather than long-horizon orchestration --- a point we use to calibrate
our results.

\subsection{Positioning}
The system most directly related to this study is the PDDL Planning Copilot
\citep{benyamin2025pddlcopilot}, which wraps planners and validators behind an MCP interface
and a single natural-language entry point. Its original evaluation was preliminary --- a
limited prompt set, no confidence intervals, and no control for memorization of public
benchmark domains. This paper contributes a substantially more rigorous evaluation of
whether such symbolic tools help LLMs: five single-tool-use PDDL tasks across five models, a
three-arm design that separates tool \emph{availability} from a one-sentence steering
\emph{nudge}, Wilson confidence intervals with signed significance, a contamination control
over anonymized domains, and a token-cost (cost-of-pass) analysis. Cross-benchmark
generalization (e.g., PlanBench \citep{valmeekam2023planbench}) and published formalizer
baselines \citep{huang2025limit} are deliberately out of scope here (see Future Work).

\section{Methodology}
% TODO (goal 3a). Sources: EXPERIMENTS_FLOW.md, run_experiment.py, pddl_eval/.
% \subsection{Tasks}                      % solve, validate_domain/problem/plan, simulate
% \subsection{Models and Serving}         % 5-model vLLM roster
% \subsection{Conditions}                 % no-tools / +tool(plain) / +tool(steered)
% \subsection{Metrics and Confidence Intervals}  % task_success_rate, tool_selected, Wilson 95%, signed-CI verdicts
% \subsection{Contamination Control}      % canonical vs anonymized domains (the null)

\section{Results}
% Structure DECIDED 2026-06-14 (regime-led hybrid) — full plan in paper/RESULTS_PLAN.md.
% Source: unified deck checkpoints/rq-sweep5v2/ + paper_notes_discussions.md. No prose yet.
% Lead paragraph + Table 1 (scorecard: 6 RQs, verdicts, headline numbers).

\subsection{Two Tasks the Model Cannot Do Without the Tool}
% RQ0.2 solve + RQ0.4 simulate (sole-source). Fig 1 = solve.png + simulate.png.

\subsection{Tasks With Headroom: The Tool Still Helps}
% RQ0.1 validate_domain + validate_problem. Text + Table 1 (charts saturate ~100%).

\subsection{When Availability Backfires, Steering Repairs It}
% RQ0.3 validate_plan (MIXED). Fig 2 = mechanism_validate_plan.png.
% Silence-not-error + signed-significance consequence + 9B>35B propensity-not-capability.

\subsection{Does the Advantage Track Difficulty?}
% RQ0.5 grows with plan length; RQ0.6 flat with object count. Mostly text; bins in appendix.

\subsection{What the Tool Costs}
% Co-headline. Fig 3 = token_quadrant.png + cost-of-pass decomposition in text.

\subsection{Robustness: Reasoning Mode and Contamination}
% Compact. think=on pattern survives + robust floor + budget caveat (Fig 4 = visible_mode_succ.png).
% Contamination null (sweep-6) in one paragraph.

\section{Limitations}
% TODO: think=on decode-budget cliff; 8192 output cap (cap-raised rerun untested);
% 5 models / 2 families; strict end-to-end grading (deliberate); latency unrecoverable;
% FastMCP relabel verified inert on sweep5v2.

\section{Future Work}
% TODO (goal 4): PlanBench results & discussion + SOTA LLM-as-formalizer baselines
% (incl. Huang & Zhang ACL 2025) are OUT OF SCOPE here — reserve ONE short paragraph.

\section{Conclusion}
% TODO.

% If AAAI-27 requires the reproducibility checklist in-paper, inline it here
% (final submission must be a SINGLE .tex file — no \input):
% \input{authorkit27/ReproducibilityChecklist.tex}

\bibliography{refs}

\end{document}
